---
title: "Statistical Analysis"
output: 
  html_document:
    toc: true
    toc_float: true 
    code_folding: hide 
editor_options: 
  chunk_output_type: console
---


This file contains ANOVA and respective post-hoc tests of the dataset.


Load & Clean Data

```{r message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(janitor)
library(lubridate)
library(leaflet)
library(crosstalk)
library(htmltools)
library(DT)
library(modelr) 
library(mgcv) 

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
  )


```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
airplane_df = read_csv("datasets/airplane_crashes_data.csv") |>
  clean_names() |>
  filter(ground != "NULL", aboard != "NULL") |>
  select(-flight_number, -time, -registration)
```

```{r, echo=FALSE}
airplane_df = airplane_df |> 
  mutate(
  
    date = str_trim(date),
    
      m = as.numeric(sub("/.*", "", date)),                    
    d = as.numeric(sub(".*/(.*)/.*", "\\1", date)),          
    y = as.numeric(sub(".*/(.*)$", "\\1", date)),            
    
    y = ifelse(y < 100, y + 1900, y),
    
    date_clean = paste(m, d, y, sep = "/"),
    
    date = mdy(date_clean),
    
    year = year(date),
    month = month(date),
    month_name = factor(month(date, 
                              label = TRUE, 
                              abbr = TRUE), 
                              levels = month.abb)
  ) |>
  select(-m, -d, -y, -date_clean) 
```

```{r, echo=FALSE}
airplane_df = airplane_df |>
  mutate(
    aboard     = as.numeric(aboard),
    fatalities = as.numeric(fatalities),
    ground     = as.numeric(ground),
    operator   = as.factor(operator)
  )
```

```{r echo = FALSE}
airplane_df = airplane_df |> 
  mutate(
    decade = floor(year / 10) * 10, 
    decade = paste0(decade, "s")
  ) |> 
  select(date, year, decade, month, month_name, everything()) 

geocoded_locations = read_csv("datasets/geocoded_cache.csv")

df_map <- airplane_df |>
  left_join(geocoded_locations, by = c("location" = "location_clean")) |>
  filter(!is.na(Latitude), !is.na(Longitude))
```
--- 

## Are total airplane crashes evenly distributed across the four seasons?

To investigate potential seasonal trends in aviation safety, we select data from northern hemisphere grouped the crash data by month of the accident. We then performed a Chi-Square Goodness of Fit Test to determine if total historical crashes are evenly distributed throughout the year. The null hypothesis assumes that accidents occur with equal frequency across all months. 

The analysis reveals a statistically significant difference in crash frequency across seasons. The Chi-Square test yields a p-value of approximately 0.0003, which is well below the 0.05 significance threshold, leading us to reject the null.

```{r chisq for crash distribution across season}

airplane_north = df_map |>
  filter(aboard > 0, !is.na(fatalities))|>
   filter(Latitude> 0)|> 
  mutate(
    survival_rate = (aboard - fatalities) / aboard,
    month_num     = month(date),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    )
  ) |>
  filter(!is.na(season)) |>
  mutate(season = fct_relevel(season, "Spring", "Summer", "Fall", "Winter"))

  
month_summary = airplane_north |>
   filter(Latitude> 0)|> 
  count(month, name = "n")|>
  # ---- compute expected + chisq ----
    mutate(
    expected = mean (n),
    chisq = list(chisq.test(n)),
    residual = n - expected,
    std_resid = (n - expected) / sqrt(expected)
  )|> print()
  
# extract p-value
chisq_p = month_summary$chisq[[1]]$p.value
chisq_p


# ---- plot ----
month_summary |>
  ggplot(aes(month, n)) +
  geom_col(fill = "steelblue") +
  geom_hline(aes(yintercept = expected),
             linetype = "dashed", size = 1) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Total Crashes by month",
    subtitle = paste("Chi-Square p-value =", format.pval(chisq_p)),
    x = "Month",
    y = "Total Number of Crashes"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Does the average yearly crashes differ across the months?
```{r anova for seaonal crashes per year}

cat("=== ANOVA result ===\n")
airplane_north |> 
  # Count crashes per year/season combo
  count(year, month, name = "crash_count")|>

# ANOVA 
aov(crash_count ~ factor(month), data = _) |> 
  summary()

# ---- plot ----
 airplane_north |> 
  count(year, month, name = "crash_count")|>
  group_by(month) |>
  summarise(
    mean_crashes = mean(crash_count),
    se = sd(crash_count) / sqrt(n())
  ) |>
  ggplot(aes(x = month, y = mean_crashes)) +
  geom_col(fill = "steelblue",show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean_crashes - se,
                    ymax = mean_crashes + se),
                width = 0.15) +
  labs(
    title = "Average Yearly Airplane Crashes in Northern Hemisphere by Month",
        x = "Month",
    y = "Average Crash Count"
  ) +
  theme_minimal()

```

## What about seasonale differences of fatality rate?

No significant evidence suggest that there are seasonal differences of fatality rate.

```{r anova for season and fatality}

cat("=== ANOVA result ===\n")

airplane_season|>
 ##ANOVA
aov(survival_rate ~ season, data = _ )|>
summary()


```
## Is the fatality rate lower over the years?

Statistical evidence confirms that airplanes have become safer over time regarding survivability. Linear regression analysis reveals a statistically significant negative correlation between year and fatality rate (p < 0.001). 
However, the low R-squared value (0.004) indicates that while the downward trend is consistent, the outcome of any individual crash is heavily influenced by situational factors other than just the time period.

```{r lm year ~ fatality}

 airplane_df |> 
  filter(aboard > 0, !is.na(fatalities)) |> 
  mutate(fatality_rate = fatalities / aboard) |> 
  group_by(year) |> 
  summarise(mean_fatality_rate = mean(fatality_rate)) |> 
  ggplot(aes(x = year, y = mean_fatality_rate)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", fill = "pink",se = TRUE) +
  labs(
    title = "Trend in Airplane Fatality Rate Over Time",,
    x = "Year",
    y = "Average Fatality Rate"
  ) +
  theme_minimal()

 airplane_df|>
  filter(aboard > 0, !is.na(fatalities)) |> 
  
  ## Calculate fatality rate
  mutate(fatality_rate = fatalities / aboard) |> 
  ## for individual event
 lm(fatality_rate ~ year, data = _)|> summary()


```


## Are the number of crash events higher or lower over the years?
```{r}

crashes_by_year = airplane_df |> 
  count(year, name = "total_crashes")

ggplot(crashes_by_year, aes(x = year, y = total_crashes)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", fill = "pink") +
  labs(
    title = "Trend of Total Airplane Crashes per Year",
    x = "Year",
    y = "Total Number of Crashes"
  ) +
  theme_minimal()

fit_count <- lm(total_crashes ~ year, data = crashes_by_year)

summary(fit_count)
```

## Cross validation : linear vs pwl vs smooth
```{r cross validation}

crashes_by_year     = airplane_df |> 
  count(year, name  = "total_crashes")|>
  mutate(year_cp1990= (year > 1990) * (year - 1990))

set.seed(1)
cv_df = crossv_mc(crashes_by_year, 100)

linear_mod = lm(total_crashes ~ year, data = crashes_by_year)
pwl_mod    = lm(total_crashes ~ year + year_cp1990, data = crashes_by_year)
smooth_mod = gam(total_crashes ~ s(year), data = crashes_by_year)
## compare 3 models
## visually smooth mod is better
crashes_by_year |> 
  gather_predictions(linear_mod, pwl_mod, smooth_mod) |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = year, y = total_crashes)) + 
  geom_point(alpha = .5) +
  geom_line(aes(y = pred), color = "red") + 
  facet_grid(~model)
             
cv_df <- cv_df |> 
  mutate(
    train = map(train, as_tibble),
    test  = map(test, as_tibble)
  ) |> 
  mutate(
    
    linear_mod = map(train, \(df) lm(total_crashes ~ year, data = df)),
    pwl_mod     = map(train, \(df) lm(total_crashes ~ year + year_cp1990, data = df)),
    smooth_mod = map(train, \(df) gam(total_crashes ~ s(year), data = df))
  ) |> 
  mutate(
    # calculate RMSE
    rmse_linear = map2_dbl(linear_mod, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_pwl   = map2_dbl(pwl_mod, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_smooth = map2_dbl(smooth_mod, test, \(mod, df) rmse(model = mod, data = df))
  )


cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_"
  ) |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + 
  geom_violin()

summary(smooth_mod)