---
title: "Descriptive Statistics and Plots"
output: 
  html_document:
    toc: true
    toc_float: true 
    code_folding: hide 
editor_options: 
  chunk_output_type: console
---

This file contains descriptive statistics plots and includes the data cleaning process involved with it.


# Data Cleaning

```{r message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(lubridate)
library(stringr)
library(plotly)
library(ggthemes)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
  )

theme_set(theme_minimal() + theme(legend.position = "bottom"))

```

If you would like to see some of our data cleaning methods, click "show" to show the code corresponding to each step of the process below: 

1. Loading and cleaning the data

```{r}

airplane_df = read_csv("datasets/airplane_crashes_data.csv") |> 
  janitor::clean_names() |> 
  filter(ground != "NULL", aboard != "NULL") |> 

  # removes unnecessary columns for our analyses
  select(-flight_number, -time, - registration)

```

2. Converting 2 digit years to four digit years

```{r}
airplane_df = airplane_df |> 
  mutate(
    # remove leading and trailing spaces
    date = str_trim(date),
    
    # extract month, day, and year from the date string
    m = as.numeric(sub("/.*", "", date)),                    
    d = as.numeric(sub(".*/(.*)/.*", "\\1", date)),          
    y = as.numeric(sub(".*/(.*)$", "\\1", date)),            
    
    # convert 2-digit years (<100) to 4-digit (1900s)
    y = ifelse(y < 100, y + 1900, y),
    
    # rebuild the string
    date_clean = paste(m, d, y, sep = "/"),
    
    # convert to Date type
    date = mdy(date_clean),
    
    # extract numeric year and month
    year = year(date),
    month = month(date),
    month_name = factor(month(date, 
                              label = TRUE, 
                              abbr = TRUE), 
                              levels = month.abb)
  ) |>
  select(-m, -d, -y, -date_clean)  # remove unnecessary columns

```

3. Converting variables to their proper variable types

```{r}

airplane_df = airplane_df |> 
  mutate(
    aboard     = as.numeric(aboard),
    fatalities =  as.numeric(ground),
    operator   = as.factor(operator)  # to group by operator
    )

```

4. Creating a decade column, now that year is numeric

```{r}
airplane_df = airplane_df |> 
  mutate(
    decade = floor(year / 10) * 10, 
    decade = paste0(decade, "s")
  ) |> 
  select(date, year, decade, month, month_name, everything()) 
```
# Descriptive Statistics

```{r}
airplane_df |>
  skimr::skim()
```

Our dataset, `airplane_df`, has 5,236 rows and 14 variables (columns). Some variables had missing values, such as `location`, `route`, `type`, `cn_in`, `summary`, and `operator`, but our key numeric variables had no missing values. p0, p50, p75, and p100 represent the minimal value, the 25th percentile, median, the 75th percentile, and the maximum value, respectively. 

# Final Plots

## Airplane Crashes Per Year
```{r, fig.width=8, fig.height=5}

airplane_df |> 
  group_by(year) |> 
  summarize(total_crashes = n()) |> 
  ggplot(aes(x = year, y = total_crashes)) + 
  geom_line(color = "blue") +
  geom_point(color = "darkblue") + 
  labs(
    title = "Airplane Crashes per Year",
       x  = "Year",
       y  = "Number of Crashes"
    ) +
  theme_fivethirtyeight() +
 theme(
    axis.title = element_text(size = 14),  # keep axis titles
    axis.text  = element_text(size = 12)   # keep axis tick labels
  )
```

In the early years of flight, there were not as many airplanes in the sky, so the relatively low number of crashes from the early 1900s to 1940 make sense. Beginning in the late 1940s to early 50's, more people started flying, which coincides with an increase in crashes. The commercial jet service was introduced in 1952, which is where we see a steady increase in crashes after that.

## Average Number of People on Planes per Year

```{r, fig.width=8, fig.height=5}
airplane_df |> 
  group_by(year) |> 
  summarize(avg_aboard = mean(aboard, na.rm = TRUE), .groups = "drop") |> 
  ggplot(aes(x = year, y = avg_aboard)) +
  geom_line(color = "darkorange", size = 1) +
  geom_point(color = "orange", size = 2) +
  labs(
    title = "Average Number of People on Planes per Year",
    x     = "Year",
    y     = "Average Number of People"
  ) +
  theme_fivethirtyeight() +
  theme(
    axis.title = element_text(size = 14),  # keep axis titles
    axis.text  = element_text(size = 12)   # keep axis tick labels
  )
```

It is useful to see the general trends in the number of people on planes per year. Overall, there is a steady upward trend in the number of people on planes as air travel became more widely tested and popularity grew.

## Top 15 Airline Operators with the Most Crashes

```{r, fig.width=8, fig.height=5}

airplane_df |> 
  group_by(operator) |> 
  summarise(total_crashes = n(), .groups = "drop") |> 
  slice_max(total_crashes, n = 15) |>  # select the top 15 operators
  ggplot(aes(x = reorder(operator, total_crashes), y = total_crashes )) +
  geom_col(fill = "blue") +
  coord_flip() +
  labs(
    title = "Top 15 Airline Operators with Most Crashes",
    x     = "Airline Operators",
    y     = "Number of Crashes"
  )  + 
  theme_fivethirtyeight() +
  theme(
    axis.title = element_text(size = 14)
  ) 
```

Aeroflot, the largest airline in Russia, is by far the commercial airline with the greatest crashes at `r airplane_df |>  filter(grepl("Aeroflot", operator, ignore.case = TRUE)) |> nrow()` total crashes. The U.S. Air Force, a military airline, comes in close second at `r airplane_df |> filter(grepl("Military - U.S. Air Force", operator, ignore.case = TRUE)) |> nrow()` total crashes.

## Aircrafts with the Most Crashes 

```{r, fig.width=8, fig.height=5}
airplane_df |> 
  filter(!is.na(type)) |> 
  group_by(type) |> 
  summarize(total_crashes = n(), .groups = "drop") |> 
  slice_max(total_crashes, n = 15) |> 
  ggplot(aes(x = reorder(type, total_crashes), y = total_crashes)) +
  geom_col(fill = "darkgreen") + 
  coord_flip() +
  labs(
    title = "Aircrafts with the Most Crashes",
    x     = "Number of Crashes",
    y     = "Aircraft Type"
  ) +  
  theme_fivethirtyeight() +
  theme(
    axis.title = element_text(size = 14)
  ) 
```

The Douglas DC-3 has had by far the most crashes out of all the aircraft types. Most of the planes in this graph are Douglas aircrafts. Interestingly, the Douglas DC-3 became the most successful airliner in the formative years of air transformation, and was the first to fly profitably without government subsidy. 

## Seasonal and Yearly Trends displayed on a Heat Map

```{r, fig.width=8, fig.height=5}

# heatmap of crashes by year and month
airplane_df |> 
  count(decade, month_name) |>  # counts crashes in the decade/month
  ggplot(aes(x = decade, y = month_name, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Heatmap of Airplane Crashes by Year and Month",
       x = "Deacde",
       y = "Month",
       fill = "Number of Crashes")

```

Based on this map, the 1970s, particularly the January, December, October, September, and August months have especially high numbers of crashes compared to other decades and months. July in the 1990s, also had a particularly high number of crashes, indicated by the darker red color. The monthly and seasonal trends will be looked at in more detail using ANOVA <a href="anova.html" target="_blank">here.</a>
